Aspects of RDF Data Constraints in the Social, Behavioural, and Economic Sciences

### 2015-07-21 paper/IEEE-ICSC-2016/IEEE-ICSC-2016.tex
classification

%All properties, not having the same domain and range types, are defined to be pairwise disjoint
%(\emph{R-9: disjoint properties}), i.e., no individual \emph{x} can be connected to an individual/literal \emph{y} by disjoint properties (e.g., \emph{phdd:isStructuredBy} and \emph{phdd:column}).
%All \emph{PHDD} classes (e.g., \emph{phdd:TableDescription}, \emph{phdd:ColumnDescription}) are pairwise disjoint (\emph{R-7: disjoint classes}),
%i.e., individuals cannot be instances of multiple disjoint classes.
\textbf{\emph{RDFS/OWL Based} constraint types.}
\textcolor{red}{severity levels assigned for each constraint?}
It is a common requirement to narrow down the value space of properties by an exhaustive enumeration of valid values (\emph{R-30/37: allowed values}):
%\emph{Allowed values} (\emph{R-30, R-37}) for properties can be IRIs (matching one or multiple patterns), any literals, allowed literals (e.g. 'red' 'blue' 'green'), and typed literals of one or multiple type(s) (e.g. \emph{xsd:string}). 
\emph{DDI-RDF:CategoryStatistics}, e.g., can only have \emph{DDI-RDF:computationBase} relationships to the values \emph{valid} and \emph{invalid} of the datatype \emph{rdf:langString} (default severity level: \emph{error}).
%Validation should \emph{exploit sub-super relations} in vocabularies (\emph{R-224}).
%If \emph{dcterms:coverage} and one of its sub properties (\emph{dcterms:spatial}, \emph{dcterms:temporal}) are given,
%it is checked that \emph{dcterms:coverage} is not redundant with its sub-properties 
%which may indicate when the data is verbose/redundant or expressed at a too general level.
Consider the following \emph{DL knowledge base} $\mathcal{K}$\footnote{A \emph{DL knowledge base} is a collection of formal statements which corresponds to \emph{facts} or what is known explicitly. For simplicity reasons, we do not state namespace prefixes.}:
\begin{center}
\begin{DL} 
$\mathcal{K}=\{$ 
	%&\ms{isStructuredBy $\sqsubseteq$ $\neg$ column}, \\
	%&\ms{TableDescription $\sqcap$ ColumnDescription $\sqsubseteq$ $\perp$}, \\
&{\small\ms{CategoryStatistics $\equiv$ $\forall$ computationBase.}} \\
	&{\small\ms{\{valid,invalid\} $\sqcap$ langString,}} \\
  &{\small\ms{Variable $\equiv$ $\exists$ concept.Concept,}} \\
	&{\small\ms{Catalog $\sqsubseteq$ $\forall$ dataset.Dataset,}} \\
  &{\small\ms{$\exists$ isStructuredBy.$\top$ $\sqsubseteq$ Table,}} \\
	&{\small\ms{$\top$ $\sqsubseteq$ $\forall$ belongsTo.Concept}}
\}\\ 
\end{DL}
\end{center}

@@

A \emph{universal quantification} (\emph{R-91}) contains all those individuals that are connected by a property only to individuals/literals of particular classes  or data ranges.
%Only \emph{dcat:Catalogs}, e.g., can have \emph{dcat:dataset} relationships to \emph{dcat:Datasets}:
%\begin{DL}
%Catalog $\sqsubseteq$ $\forall$ dataset.Dataset
%\end{DL}
Only \emph{dcat:Catalogs}, e.g., can have \emph{dcat:dataset} relationships to \emph{dcat:Datasets} (\emph{error}).
\emph{Property domains} (\emph{R-25, R-26}) and \emph{property ranges} (\emph{R-28, R-35}) constraints restrict domains and ranges of properties:
Only \emph{phdd:Tables}, e.g., can have \emph{phdd:isStructuredBy} relationships and
\emph{xkos:belongsTo} relations (\emph{error}) can only point to \emph{skos:Concepts} (\emph{error}).
%\emph{Existential quantifications} (\emph{R-86}) enforce that instances of given classes must have some property relation to individuals/literals of certain types.
%Variables, e.g., should have a relation to a theoretical concept ($\mathcal{SL}_{0}$).
%%The variables \emph{number of school years} and \emph{highest educational degree}, e.g., are associated with the theoretical concept \emph{education}. 
%The variable \emph{Education at pre-school} is associated with the theoretical concept \emph{Child Care}. 
%The default severity level of the constraint is weak, as in most cases research can be continued without having information about the theoretical concept of a variable.
%If a study, e.g., does not contain any data set ($\mathcal{SL}_{2}$), the actual description of the data is missing which may indicate that it is very hard or unlikely to get access to the data.
%If metadata on data files, including the actual data, is missing (especially case and variable quantities; $\mathcal{SL}_{1}$), the description of the data sets and the study is not sufficient.
%Case quantity measures how many cases are collected for a study.
%High case and variable quantities are indicators for high statistical quality and comprehensiveness of the underlying study ($\mathcal{SL}_{1}$).

@@

\textbf{\emph{Constraint Language Based} constraint types.}
%{Simple constraint types} is the set of constraint types whose constraints can be easily defined without much effort in addition to $\mathcal{CT}_{B}$ constraints.
For data properties, it may be desirable to restrict that values of predefined languages must be present for determined number of times (\emph{R-48/49: language tag cardinality}):
(1) It is checked if literal language tags are set. Some controlled vocabularies, e.g., contain literals in natural language, but without information what language has actually been used (\emph{warning}). 
(2) Language tags must conform to language standards (\emph{error}). 
(3) Some thesaurus concepts are labeled in only one, others in multiple languages. 
It may be desirable to have each concept labeled in each of the languages that are also used on the other concepts,
as language coverage incompleteness for some concepts may indicate shortcomings of thesauri (\emph{informational})
\cite{MaderHaslhoferIsaac2012}.
%\textbf{Validation and Reasoning.}

\textcolor{red}{\emph{Default values}} (\emph{R-31, R-38}) for objects/literals of given properties are inferred automatically when properties are not present in the data.
The value \emph{true} for the property {\em DDI-RDF:isPublic} indicates that a {\em DDI-RDF:LogicalDataSet} can be accessed by anyone.
Per default, however, access to data sets should be restricted (\emph{false}) (\emph{informational}).
Many properties are not necessarily required but \emph{recommended} within a particular context (\emph{R-72}).
The property {\em skos:notation}, e.g., is not mandatory for {\em DDI-RDF:Variable}s, but recommended to represent variable names (\emph{informational}).

Depending on property datatypes, two different literal values have
a specific ordering with respect to operators like \textless  (\textcolor{red}{\emph{R-43: literal value comparison}}).
Start dates (\emph{DDI-RDF:startDate}), e.g., must be before (\emph{\textless}) end dates (\emph{DDI-RDF:endDate}).

Percentage values are only valid when they are within the literal range of 0 and 100 (\emph{R-45: literal ranges}; \emph{error})
which is checked for \emph{DDI-RDF:percentage} standing for the number of cases of a given code in relation to the total number of cases for a particular variable.

@@

It is often useful to declare a given (data) property as the \emph{primary key} (\emph{R-226}) of a class, so that a system can enforce uniqueness and build URIs from user inputs and imported data. 
In DDI-RDF, resources are uniquely identified by the property \emph{adms:identifier},
which is therefore inverse-functional $({\small\ms{funct identifier}\sp{\overline{\ }}})$,
i.e., for each \emph{rdfs:Resource x}, there can be at most one distinct resource \emph{y} such that \emph{y} is connected by \emph{adms:identifier$\sp{\overline{\ }}$} to \emph{x} (\emph{error}).
%\begin{DL}
%$(\ms{funct identifier}\sp{\overline{\ }})$
%\end{DL} 
Keys, however, are even more general than \emph{inverse-functional properties} (\emph{R-58}),
as a key can be a data property, an object property, or a chain of properties \cite{Schneider2009}.
Thus and as there are different sorts of key, and as keys can lead to undecidability, 
DL is extended with the construct \emph{keyfor} ({\small\ms{identifier \ms{keyfor} Resource}}) \cite{Lutz2005} which is implemented by the \emph{OWL 2} \emph{hasKey} construct.
%\emph{OWL 2} \emph{hasKey} implements \emph{keyfor} ($\mathcal{SL}_{2}$) and thus can be used to identify resources uniquely, to merge resources with identical key property values, and to recognize constraint violations.
%\begin{DL}
%identifier \ms{keyfor} Resource
%\end{DL}

@@

In many cases, resources must be \emph{members of controlled vocabularies} (\emph{R-32}).
If a dimension property, e.g., has a \emph{qb:codeList},
then the value of the dimension property on every \emph{qb:Observation} must be in the code list (\emph{error}).
%Thus, summary statistics can only have \emph{DDI-RDF:summaryStatisticType} relationships to \emph{skos:Concept}s which must be members of the controlled vocabulary \emph{ddicv:SummaryStatisticType}, a \emph{skos:ConceptScheme} ($\mathcal{SL}_{2}$).
Objects/literals can be declared to be ordered for given properties (\emph{R-121/217: ordering}).
Variables, questions, and codes, e.g., are typically organized in a particular order. 
If codes (\emph{skos:Concept}) should be ordered, they must be members (\emph{skos:memberList}) in an ordered collection (\emph{skos:OrderedCollection}), the variable's code list (\emph{informational}).

%\textbf{Constraints on Properties.}
It is useful to declare properties to be \emph{conditional} (\emph{R-71}), i.e., if particular properties exist (or do not exist), then other properties must also be present (or absent).
To get an overview over a series/study either an abstract, a title, an alternative title, or links to external descriptions should be provided. 
If an abstract and an external description are absent, however,  
a title or an alternative title should be given (\emph{warning}).
In case a variable is represented in form of a code list, codes may be associated with categories, i.e., human-readable labels (\emph{informational}).
The variable \emph{Education at pre-school}, e.g., is represented as ordered code list without any categories.
%If a {\em skos:Concept} represents a code (having {\em skos:notation} and {\em skos:prefLabel} properties), 
%then the property {\em DDI-RDF:isValid} has to be stated indicating if the code stands for valid (\emph{true}) or missing (\emph{false}) cases ($\mathcal{SL}_{2}$).

>>>>

%All properties, not having the same domain and range types, are defined to be pairwise disjoint
%(\emph{R-9: disjoint properties}), i.e., no individual \emph{x} can be connected to an individual/literal \emph{y} by disjoint properties (e.g., \emph{phdd:isStructuredBy} and \emph{phdd:column}).
%All \emph{PHDD} classes (e.g., \emph{phdd:TableDescription}, \emph{phdd:ColumnDescription}) are pairwise disjoint (\emph{R-7: disjoint classes}),
%i.e., individuals cannot be instances of multiple disjoint classes.
\textbf{\emph{RDFS/OWL Based} constraint types.}
It is a common requirement to narrow down the value space of properties by an exhaustive enumeration of valid values (\emph{R-30/37: allowed values}):
%\emph{Allowed values} (\emph{R-30, R-37}) for properties can be IRIs (matching one or multiple patterns), any literals, allowed literals (e.g. 'red' 'blue' 'green'), and typed literals of one or multiple type(s) (e.g. \emph{xsd:string}). 
\emph{disco:CategoryStatistics}, e.g., can only have \emph{disco:computationBase} relationships to the values \emph{valid} and \emph{invalid} of the datatype \emph{rdf:langString} (default severity level: \emph{error}).
%Validation should \emph{exploit sub-super relations} in vocabularies (\emph{R-224}).
%If \emph{dcterms:coverage} and one of its sub properties (\emph{dcterms:spatial}, \emph{dcterms:temporal}) are given,
%it is checked that \emph{dcterms:coverage} is not redundant with its sub-properties 
%which may indicate when the data is verbose/redundant or expressed at a too general level.
Consider the following \emph{DL knowledge base} $\mathcal{K}$\footnote{A \emph{DL knowledge base} is a collection of formal statements which correspond to \emph{facts} or what is known explicitly. For simplicity reasons, we do not state namespace prefixes.}:
\begin{center}
\begin{DL} 
$\mathcal{K}=\{$ 
	%&\ms{isStructuredBy $\sqsubseteq$ $\neg$ column}, \\
	%&\ms{TableDescription $\sqcap$ ColumnDescription $\sqsubseteq$ $\perp$}, \\
&{\small\ms{CategoryStatistics $\equiv$ $\forall$ computationBase.}} \\
	&{\small\ms{\{valid,invalid\} $\sqcap$ langString,}} \\
  &{\small\ms{Variable $\equiv$ $\exists$ concept.Concept,}} \\
	&{\small\ms{Catalog $\sqsubseteq$ $\forall$ dataset.Dataset,}} \\
  &{\small\ms{$\exists$ isStructuredBy.$\top$ $\sqsubseteq$ Table,}} \\
	&{\small\ms{$\top$ $\sqsubseteq$ $\forall$ belongsTo.Concept}}
\}\\ 
\end{DL}
\end{center}

@@

A \emph{universal quantification} (\emph{R-91}) contains all those individuals that are connected by a property only to individuals/literals of particular classes  or data ranges.
%Only \emph{dcat:Catalogs}, e.g., can have \emph{dcat:dataset} relationships to \emph{dcat:Datasets}:
%\begin{DL}
%Catalog $\sqsubseteq$ $\forall$ dataset.Dataset
%\end{DL}
Only \emph{\textcolor{red}{dcat}:Catalogs}, e.g., can have \emph{dcat:dataset} relationships to \emph{dcat:Datasets} (\emph{error}).
\emph{Property domains} (\emph{R-25, R-26}) and \emph{property ranges} (\emph{R-28, R-35}) constraints restrict domains and ranges of properties:
Only \emph{phdd:Tables}, e.g., can have \emph{\textcolor{red}{phdd}:isStructuredBy} relationships (\emph{error}) and \emph{\textcolor{red}{xkos}:belongsTo} relations can only point to \emph{skos:Concepts} (\emph{error}).

It is often useful to declare a given (data) property as the \emph{primary key} (\emph{R-226}) of a class, so that a system can enforce uniqueness and build URIs from user inputs and imported data. 
In DDI-RDF, resources are uniquely identified by the property \emph{adms:identifier},
which is therefore inverse-functional $({\small\ms{funct identifier}\sp{\overline{\ }}})$,
i.e., for each \emph{rdfs:Resource x}, there can be at most one distinct resource \emph{y} such that \emph{y} is connected by \emph{adms:identifier$\sp{\overline{\ }}$} to \emph{x} (\emph{error}).
%\begin{DL}
%$(\ms{funct identifier}\sp{\overline{\ }})$
%\end{DL} 
Keys, however, are even more general than \emph{inverse-functional properties} (\emph{R-58}),
as a key can be a data property, an object property, or a chain of properties \cite{Schneider2009}.
Thus and as there are different sorts of key, and as keys can lead to undecidability, 
DL is extended with the construct \emph{keyfor} ({\small\ms{identifier \ms{keyfor} Resource}}) \cite{Lutz2005} which is implemented by the \emph{OWL 2} \emph{hasKey} construct.

%\emph{Existential quantifications} (\emph{R-86}) enforce that instances of given classes must have some property relation to individuals/literals of certain types.
%Variables, e.g., should have a relation to a theoretical concept ($\mathcal{SL}_{0}$).
%%The variables \emph{number of school years} and \emph{highest educational degree}, e.g., are associated with the theoretical concept \emph{education}. 
%The variable \emph{Education at pre-school} is associated with the theoretical concept \emph{Child Care}. 
%The default severity level of the constraint is weak, as in most cases research can be continued without having information about the theoretical concept of a variable.
%If a study, e.g., does not contain any data set ($\mathcal{SL}_{2}$), the actual description of the data is missing which may indicate that it is very hard or unlikely to get access to the data.
%If metadata on data files, including the actual data, is missing (especially case and variable quantities; $\mathcal{SL}_{1}$), the description of the data sets and the study is not sufficient.
%Case quantity measures how many cases are collected for a study.
%High case and variable quantities are indicators for high statistical quality and comprehensiveness of the underlying study ($\mathcal{SL}_{1}$).

@@

\textbf{\emph{Constraint Language Based} constraint types.}
%{Simple constraint types} is the set of constraint types whose constraints can be easily defined without much effort in addition to $\mathcal{CT}_{B}$ constraints.
%\textbf{Validation and Reasoning.}

Depending on property datatypes, two different literal values have
a specific ordering with respect to operators like \textless  (\emph{R-43: literal value comparison}).
Start dates (\emph{disco:startDate}), e.g., must be before (\emph{\textless}) end dates (\emph{disco:endDate}).

%Percentage values are only valid when they are within the literal range of 0 and 100 (\emph{R-45: literal ranges}; \emph{error})
%which is checked for \emph{DDI-RDF:percentage} standing for the number of cases of a given code in relation to the total number of cases for a particular variable.

In many cases, resources must be \emph{members of controlled vocabularies} (\emph{R-32}).
If a QB dimension property, e.g., has a \emph{qb:codeList},
then the value of the dimension property on every \emph{qb:Observation} must be in the code list (\emph{error}).

\emph{Default values} (\emph{R-31, R-38}) for objects/literals of given properties are inferred automatically when properties are not present in the data.
The value \emph{true} for the property {\em disco:isPublic} indicates that a {\em disco:LogicalDataSet} can be accessed by anyone.
Per default, however, access to data sets should be restricted (\emph{false}) (\emph{informational}).

%Many properties are not necessarily required but \emph{recommended} within a particular context (\emph{R-72}).
%The property {\em skos:notation}, e.g., is not mandatory for {\em disco:Variable}s, but recommended to represent variable names (\emph{informational}).

@@

%\emph{OWL 2} \emph{hasKey} implements \emph{keyfor} ($\mathcal{SL}_{2}$) and thus can be used to identify resources uniquely, to merge resources with identical key property values, and to recognize constraint violations.
%\begin{DL}
%identifier \ms{keyfor} Resource
%\end{DL}

@@

%Summary statistics types like minimum, maximum, and arithmetic mean are maintained within a controlled vocabulary.  
%Thus, summary statistics can only have \emph{DDI-RDF:summaryStatisticType} relationships to \emph{skos:Concept}s which must be members of the controlled vocabulary \emph{ddicv:SummaryStatisticType}, a \emph{skos:ConceptScheme} ($\mathcal{SL}_{2}$).
Objects/literals can be declared to be ordered for given properties (\emph{R-121/217: ordering}).
Variables, questions, and codes, e.g., are typically organized in a particular order. 
If codes (\emph{skos:Concept}) should be ordered, they must be members (\emph{skos:memberList}) in an ordered collection (\emph{skos:OrderedCollection}), the variable's code list (\emph{informational}).

%\textbf{Constraints on Properties.}
It is useful to declare properties to be \emph{conditional} (\emph{R-71}), i.e., if particular properties exist (or do not exist), then other properties must also be present (or absent).
To get an overview over a series/study either an abstract, a title, an alternative title, or links to external descriptions should be provided. 
If an abstract and an external description are absent, however,  
a title or an alternative title should be given (\emph{warning}).
In case a variable is represented in form of a code list, codes may be associated with categories, i.e., human-readable labels (\emph{informational}).
The variable \emph{Education at pre-school}, e.g., is represented as ordered code list without any categories.

For data properties, it may be desirable to restrict that values of predefined languages must be present for determined number of times (\emph{R-48/49: language tag cardinality}):
(1) It is checked if literal language tags are set. Some controlled vocabularies, e.g., contain literals in natural language, but without information what language has actually been used (\emph{warning}). 
(2) Language tags must conform to language standards (\emph{error}). 
(3) Some thesaurus concepts are labeled in only one, others in multiple languages. 
It may be desirable to have each concept labeled in each of the languages that are also used on the other concepts,
as language coverage incompleteness for some concepts may indicate shortcomings of thesauri (\emph{informational})
\cite{MaderHaslhoferIsaac2012}.

%If a {\em skos:Concept} represents a code (having {\em skos:notation} and {\em skos:prefLabel} properties), 
%then the property {\em DDI-RDF:isValid} has to be stated indicating if the code stands for valid (\emph{true}) or missing (\emph{false}) cases ($\mathcal{SL}_{2}$).
